Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-06 paper_jasa.tex
abstract updated
small changes

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against conventional statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.
\end {abstract}

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$, with an exact $x_{\alpha}$. For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case \blue{(Is it true? For conventional test it is easy to compute Type II error !)}. In visual inference, individual abilities need to be incorporated to calculate Type II error. This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. For Type II error, we need to calculate the chance that the observer sees that actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise, where $p_i$ is the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

@@

\hh{\noindent An important difference between conventional testing and the lineup scenarios \blue{(Can we just say 'visual test'?)} is that lineups always depend on observers -- this brings the number of people evaluating a particular lineup into the actual evaluation of power, i.e. we are assuming, again, that $X$ counts the number of observers (out of $K$) who identify the data plot from the lineup.
Then power is calculated as
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
where $F_{X, \theta}$ is the distribution of $X$ -- this depends on which hypothesis is true:

Under the null hypothesis $X \sim Binom_{K, 1/m}$, leading to a power of \blue{(Is this power or Type I error?)}
\[
\text{Power}_V(\theta, K)= 1 - Binom_{K, 11/m} (x_\alpha - 1),
\]
where $x_\alpha$ is defined as in (\ref{dfn:test}), as  $P(X \ge x_{\alpha}) < \alpha$.

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against conventional statistical testing  of the significance of regression model parameters. \blue{We extend and refine the terminology of visual inference to a more general framework to allow for comparisons with conventional tests.} A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.
\end {abstract}

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$, with an exact $x_{\alpha}$. For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. For Type II error, we need to calculate the chance that the observer sees that actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise, where $p_i$ is the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

@@

\hh{\noindent An important difference between conventional and visual testing  is that lineups always depend on observers -- this brings the number of people evaluating a particular lineup into the actual evaluation of power, i.e. we are assuming, again, that $X$ counts the number of observers (out of $K$) who identify the data plot from the lineup.
Then power is calculated as
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
where $F_{X, \theta}$ is the distribution of $X$ -- this depends on which hypothesis is true:

Under the null hypothesis $X \sim Binom_{K, 1/m}$, leading to a type I error of:
\[
\text{Power}_V(\theta, K)= 1 - Binom_{K, 11/m} (x_\alpha - 1),
\]
where $x_\alpha$ is defined as in (\ref{dfn:test}), as  $P(X \ge x_{\alpha}) < \alpha$.
